---
title: "One day, one country"
---

# Project description

Every day I randomly pick a new country, then I read and write a bit about it. The idea is to learn something new every day about the world and get into the habit of writing regularly. As I write about a new country every day, I don't have time to research things carefully. This means there may be many inaccuracies and outright errors. Please feel free to point them out to me if you encounter any. I primarily think of these notes as a public diary in which I keep track in (almost) real-time of the things I found interesting and the questions I had as I was learning about a new country.

# Overview

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r prepare-table}
posts <- tidyr::tribble(
  ~Day, ~Country, ~Continent, ~Notes, 
  "Day 7", "Seychelles", "Africa", "[See notes for Seychelles](#seychelles)",
  "Day 6", "Malta", "Europe", "[See notes for Malta](#malta)",
  "Day 5", "Uzbekistan", "Asia", "[See notes for Uzbekistan](#uzbekistan)",
  "Day 4", "Suriname", "South America", "[See notes for Suriname](#suriname)",
  "Day 3", "Burundi", "Africa", "[See notes for Burundi](#burundi)",
  "Day 2", "Uganda", "Africa", "[See notes for Uganda](#uganda)",
  "Day 1", "Kenya", "Africa", "[See notes for Kenya](#kenya)"

)
```

```{r display-table}
knitr::kable(posts)
```

# Notes

## Seychelles

For a small country of around 100 thousand inhabitants, one wouldn’t expect a lot of political turmoil. We would be wrong in this case. Since Seychelles gained its independence from the United Kingdom in 1976, it saw, among other things, a successful coup d’état, various failed ones, a one-party state, the murder of an opposition leader in London, and a peaceful transition of power. 

One year into its independence, the country’s elected president was deposed while attending a conference outside the country. In a virtually bloodless coup, around 200 armed men took over strategic positions on Mahé, the main island. Based on a contemporary [New York Times article]( https://www.nytimes.com/1977/06/07/archives/seychelles-capital-is-calm-after-coup-insurgents-secretive-about.html) that describes the aftermath of the events, one comes away with the impression that the inhabitants’ lives hadn’t been upended in significant ways. For example, the article notes that tourists on the island were a bit annoyed they couldn’t buy alcohol temporarily and that six British officers were arrested and flown back to Europe. After this coup, the person who served as the prime minister under the deposed president formed a new government and quickly instituted a one-party state.

Four years later, in 1981, the country saw an unsuccessful coup attempt, led by an Irish mercenary and orchestrated by South Africa. A group of 44 mercenaries flew into Mahé, under the pretext that they were rugby players on holiday. Their identity was quickly discovered when a baggage search produced an AK-47, and the coup attempt failed. 

In 1991, the same president who introduced the one-party state announced a return to a multiparty system. He won several elections before voluntarily turning over his presidency to another politician from his party. In 2020, the elections were won by the opposition party that hadn’t been in power since 1977.

In sum, Seychellois politics of the last couple of decades produced many rather unexpected (and occasionally bizarre) episodes.


## Malta

Maltese is one of the two official languages in Malta. It is a Semitic language, making Malta the only country in the European Union to use one. How did Maltese become Malta’s official language? To answer this question, we need to take a step back and briefly consider Malta’s geographic location and history.

The most widely spoken Semitic language is Arabic. The Arab world is geographically close (around 300 km) to Malta, with Tunisia to its west and Libya to its south. By the beginning of the 10th century, the Arabs conquered Sicily, located to the north of Malta, and created the Emirate of Sicily, which also included Malta. The conquerors spoke a variety of Arabic called Siculo-Arabic. This language is now extinct, but it served as the starting point for the evolution of Maltese.

In 1091, Norman forces conquered Malta, and the island’s re-Christianisation began. The language went through a process of Latinisation: the formerly spoken language was written down in Latin script. It’s the only Semitic language written exclusively with Latin characters, albeit it has some extra, diacritically altered letters such as ċ. Over the centuries, Italian and Sicilian had a large influence over it. As a result, its vocabulary does not only contain words of Arabic origin but also many loanwords from these two languages and, more recently, English. It has been Malta’s official language since 1934, and around 97% of the population consider it their mother tongue.


## Uzbekistan

Historically, cotton has played an important role in Uzbekistan. For instance, all the emblems the country had in the last roughly 100 years featured a cotton plant. Cotton production greatly increased during the Soviet era: around 70% of the Soviet Union’s entire production came from the Uzbek SSR. Although cotton production is now lower than it was in the Soviet era, Uzbekistan is still among the largest producers and exporters of cotton. Nicknamed as “white gold”, cotton has undoubtedly contributed to the country’s economy, but its production also brought with it some problems. Let’s briefly mention three of them.

First, cotton is a water-intensive crop. The water for its irrigation was supplied in part by the Aral Sea. Furthermore, rivers feeding the Aral Sea (Amu Darya and Syr Darya) were diverted for Soviet irrigation projects. As a result of all this, the Aral Sea started shrinking in the 1960s and lost around 90% of its original size, creating the world’s most recent desert (Aralkum Desert) in the dried-up basin.

Second, the environmental damage was exacerbated by the extensive use of fertilizers and other chemical products, which caused soil contamination and public health problems. For example, a defoliant called Butifos (since outlawed) was blamed by locals for [higher than usual infant mortality and birth defects]( https://www.newscientist.com/article/mg12416910-800-soviet-cotton-threatens-a-regions-sea-and-its-children/). 

Finally, some reports claim that cotton production in Uzbekistan is rife with human rights abuses. According to a 2020 [Bureau of International Labor Affairs report](https://www.dol.gov/agencies/ilab/reports/child-labor/list-of-goods), forced labor is still used, and some earlier reports documented child labor as well.


## Suriname

Suriname has enormous ethnic diversity for a country of only 600 thousand people.  To understand why this is the case, let’s briefly look at some of its history. Some areas of today’s Suriname were inhabited by indigenous tribes, the largest one of which were the Arawak. When Suriname became a Dutch colony in the second half of the 17th century, new plantations could be created in the fertile Guiana plains, which required a lot of human labor. Initially, this labor was provided by slaves from sub-Saharan Africa. Many of them managed to escape from the plantations and find refuge in the adjacent rainforest. The descendants of these people are the Maroons and constitute around 22% of the Surinamese.

After the Dutch abolished slavery in 1863, the labor-intensive plantations had to find another source of labor. Contract and indentured (i.e., people working without a salary for some years) workers were brought in primarily from today’s Indonesia and India. Like Suriname, Indonesia was also a Dutch colony, which made transporting people to Suriname easier. Organizing the transport of Indian workers was a bit more involved because, India being under British colonial rule, arrangements needed to be made with the British. Today around 15% of the population have Javanese (from Java in Indonesia) ancestry and around 27% have Indian ancestry. We also find people with Creole, Chinese, and Brazilian heritage among the country’s residents. 

To sum it up, Suriname’s current ethnic composition offers a glimpse of its history: the slave trade from Africa, contract and indentured workers from Asia, and recent voluntary migration from places such as Brazil. 


## Burundi

Countries in the same region tend to have many things in common. One of the aspects in which Burundi is similar to some other countries in the region (Tanzania, in particular, comes to mind) is that it has recently changed its capital city. Historically, the capital of the pre-colonial Kingdom of Burundi was Gitega, which is now the second-largest city in Burundi. Around the time Burundi gained its independence from Belgium, Bujumbura became its capital city. Now it’s the country’s largest and economically most important city, conveniently located on the shore of Lake Tanganyika. A couple of years ago, Burundi’s then-president announced that the capital would be Gitega again.

I’ve always wondered why countries change their capitals. I assume sometimes there is a purely political reason, such as moving to a politically more welcoming region. Other times, changing the capital can reflect, among other things, a genuine shift in the importance of certain regions within a country.

Finally, let me mention two things briefly in which Burundi is a bit atypical in sub-Saharan Africa. First, Burundi is linguistically more homogeneous than most other countries in the region: almost everyone speaks Kirundi, while in other countries there is usually a large degree of linguistic diversity. Second, the current country’s territory occupies that of the pre-colonial Kingdom of Burundi. This is rare in that most current borders in the region don’t coincide with those of pre-colonial states. 


## Uganda

Uganda is named after Buganda, its largest region. Buganda is in the southern part of the
country, and approximately 25% of the country's 45 million residents live there---many
of them in the capital, Kampala. People from Buganda are collectively called
the Baganda (or Ganda), while a single person from this group is a Muganda. They
speak a language called Luganda.

In yesterday's post about Kenya, I mentioned that flags may contain useful information about a
country. So, I duly checked Uganda's flag. It depicts a grey crowned crane,
Uganda's national bird. So far so good. Today's revelation (or reminder,
rather) is that names can also contain interesting pieces of information.

Suspecting that Buganda, Luganda, and Baganda may have something to do with the country's name didn't require a lot of effort because the words are almost identical. Without knowing
too much about the country's history, one would guess that the Baganda people must have played an important role in shaping Uganda's past and
present.

Detecting that some words share a common root is not always as straightforward as in
Uganda's case. Had the words been a bit more different from one
another, I could have easily overlooked the link between the country's name and its most important region.

## Kenya

I've heard about the Maasai before, but I haven't realized they are one of the largest ethnic groups in Kenya. It turns out that Kenya's flag even features a Maasai shield. A note to my future self: checking out the symbols on a flag can give me some interesting leads. Another ethnic group I found interesting are the Kalenjin: many of the famous Kenyan middle- and long-distance runners are from this ethnic group. It would be interesting to know why.

I also learned that the largest urban slum in Africa is a place called Kibera, on the outskirts of Nairobi, Kenya's capital. My questions: (1) Why there? In other words, is there a geographic, legal, or other reason why the slum is there, or is its location just the result of a random process?; (2) Did its residents move there from a particular region in Kenya (or East Africa)?

Ngũgĩ wa Thiong'o is one of the most popular Kenyan writers. He has a short story (*The Upright Revolution: Or Why Humans Walk Upright*) that has been translated into 100 languages. Maybe it's time for me to read some Kenyan literature.

While I was reading about Kenya, I noticed many names and concepts that start with the letter K---for example, Kenya itself, or the aforementioned Kalenjin and Kibera. Is this a coincidence, or is the letter K indeed common in some or all the languages Kenyans use?
