---
title: "One day, one country"
---

# Project description

Every day I randomly pick a new country, then I read and write a bit about it. The idea is to learn something new every day about the world and get into the habit of writing regularly. As I write about a new country every day, I don't have time to research things carefully. This means there may be many inaccuracies and outright errors. Please feel free to point them out to me if you encounter any. I primarily think of these notes as a public diary in which I keep track in (almost) real-time of the things I found interesting and the questions I had as I was learning about a new country.

# Overview

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r prepare-table}
posts <- tidyr::tribble(
  ~Day, ~Country, ~Continent, ~Notes,
  "Day 34", "Malawi", "Africa", "[See notes for Malawi](#malawi)",
  "Day 33", "El Salvador", "America", "[See notes for El Salvador](#el-salvador)",
  "Day 32", "Cyprus", "Europe", "[See notes for Cyprus](#cyprus)",
  "Day 31", "Panama", "America", "[See notes for Panama](#panama)",
  "Day 30", "Niger", "Africa", "[See notes for Niger](#niger)",
  "Day 29", "Costa Rica", "America", "[See notes for Costa Rica](#costa-rica)",
  "Day 28", "Haiti", "America", "[See notes for Haiti](#haiti)",
  "Day 27", "Syria", "Asia", "[See notes for Syria](#syria)",
  "Day 26", "Nigeria", "Africa", "[See notes for Nigeria](#nigeria)",
  "Day 25", "Central African Republic", "Africa", "[See notes for the Central African Republic](#central-african-republic)",
  "Day 24", "Italy", "Europe", "[See notes for Italy](#italy)",
  "Day 23", "Brazil", "South America", "[See notes for Brazil](#brazil)",
  "Day 22", "Jordan", "Asia", "[See notes for Jordan](#jordan)",
  "Day 21", "Libya", "Africa", "[See notes for Libya](#libya)",
  "Day 20", "Mauritania", "Africa", "[See notes for Mauritania](#mauritania)",
  "Day 19", "Sri Lanka", "Asia", "[See notes for Sri Lanka](#sri-lanka)",
  "Day 18", "Guyana", "South America", "[See notes for Guyana](#guyana)",
  "Day 17", "Benin", "Africa", "[See notes for Benin](#benin)",
  "Day 16", "Namibia", "Africa", "[See notes for Namibia](#namibia)",
  "Day 15", "Ivory Coast", "Africa", "[See notes for Ivory Coast](#ivory-coast)",
  "Day 14", "Ghana", "Africa", "[See notes for Ghana](#ghana)",
  "Day 13", "Gabon", "Africa", "[See notes for Gabon](#gabon)",
  "Day 12", "Kyrgyzstan", "Asia", "[See notes for Kyrgyzstan](#kyrgyzstan)",
  "Day 11", "Chad", "Africa", "[See notes for Chad](#chad)",
  "Day 10", "Lesotho", "Africa", "[See notes for Lesotho](#lesotho)",
  "Day 9", "South Sudan", "Africa", "[See notes for South Sudan](#south-sudan)",
  "Day 8", "Germany", "Europe", "[See notes for Germany](#germany)",
  "Day 7", "Seychelles", "Africa", "[See notes for Seychelles](#seychelles)",
  "Day 6", "Malta", "Europe", "[See notes for Malta](#malta)",
  "Day 5", "Uzbekistan", "Asia", "[See notes for Uzbekistan](#uzbekistan)",
  "Day 4", "Suriname", "South America", "[See notes for Suriname](#suriname)",
  "Day 3", "Burundi", "Africa", "[See notes for Burundi](#burundi)",
  "Day 2", "Uganda", "Africa", "[See notes for Uganda](#uganda)",
  "Day 1", "Kenya", "Africa", "[See notes for Kenya](#kenya)"

)
```

```{r display-table}
knitr::kable(posts)
```

# Notes

## Malawi

Since its independence in 1964, Malawi (formerly Nyasaland) was ruled by one man for three decades. The ruler, Hastings Banda, was born in the central part of the country on an unknown date, sometime around the turn of the 20th century. He spent time around various missionaries in his childhood and early adulthood. These religious connections opened doors for him; he studied in the U.S. and the U.K., then became a practicing doctor in the U.K. 

Some years after his return to Malawi, he became the country’s first leader upon its independence. He built an autocratic rule while maintaining a pro-Western, anti-communist stance. As is usually the case, he gradually tried to seize ever more power. First, he changed the constitution and got himself elected as the country’s president for a 5-year term. Although Malawi had been a de facto one-party state by then, he banned all parties but his---just in case. To save the hassles of being re-elected every five years, he was declared president for life before the end of his first term.

In many respects, Malawi functioned as a police state. Political dissent was not tolerated, censorship was rampant, and citizens frequently reported each other. In 1983, in what became known as the Mwanza Four incident, four high-ranking government officials died under suspicious circumstances, not long after they had criticized some aspects of the regime. Many years later, no longer the president, he was charged in this case (and acquitted due to lack of evidence). 

In 1993, a referendum asked Malawians if they want a one-party or a multi-party state. The multi-party side won with 65% of the votes. Put differently, 35% of the voting population favored the continuation of the existing system---a result worth thinking about. Banda contested the position of president in the country’s first multi-party elections a year later, losing to his opponent.  

## El Salvador

People all over the world use money to exchange goods and services. Money can manifest itself in many forms---anything works as money as long as you can use it to get, for example, a loaf of bread or a haircut. Before El Salvador’s colonial period, cacao beans fulfilled this important function. When the Spanish colonialists arrived, they brought their currency, the real, with them. In reality, however, the most used coins were the so-called macacos, irregularly shaped silver coins from Peru and Mexico. 

El Salvador became independent in the 19th century. In 1883, it introduced the peso. Less than ten years later, it changed its name to colón, to commemorate Christopher Columbus (called Colón in Spanish). The currency’s value was linked to the U.S. dollar through a fixed exchange rate. In the 1930s, the fixed exchange rate was abandoned.

The colón was the main currency until 2001 when a second currency was also added as legal tender: the U.S. dollar. Once the decision had been made to abandon its own currency, adopting the dollar was the most natural choice for El Salvador. Although surely many factors played a role in this decision, two are worth highlighting. First, the U.S. is the country’s most important trading partner. Second, many Salvadorians live in the U.S. and regularly send money to their families in El Salvador.

The latest development in the monetary history of El Salvador is its embrace of the cryptocurrency Bitcoin. In a move that surprised many, El Salvador’s current president, Nayib Bukele, made it legal tender last year. 

## Cyprus

Cyprus is an island in the Mediterranean Sea, located to the south of Turkey. As an island with a strategic location, it has drawn the attention of regional powers throughout its history. More than two millennia ago, Alexander the Great conquered it and Hellenized it (i.e., made the local population adopt the Greek language and culture). Fast forward many years and different rulers, the island came under the rule of the Ottoman Empire in the 16th century. People from nearby Anatolia (the peninsula that comprises a large part of today’s Turkey) settled there. As a result of the two conquests, the island came to have a population of both Greek and Turkish Cypriots.

After the Ottoman rule, the island became a British colony before finally achieving independence in 1960. The now-independent country faced a problem: its Greek and Turkish citizens wanted different things. Many Greeks, who made up around 80% of the island’s population, wanted a union with Greece---something Turkish Cypriots opposed. After bouts of intercommunal violence, a Greece-led coup d’état removed the island’s sitting president. Five days later, the Turkish army invaded the island, occupying its north-eastern part. 

The territorial division remains to this day. At the time of the Turkish invasion, many Greek Cypriots lived on the occupied side. Similarly, many Turkish Cypriots lived on the Greek side. Those Greek and Turkish Cypriots who found themselves on the wrong side of the new de-facto border were forced to move. A wall was erected, and a UN buffer zone was created between the two sides. Although the border wall was demolished in 2008, and traveling between the sides became easier, reunification efforts have failed so far. 

## Panama

Panama has a unique geographical position: it’s on a narrow strip of land connecting South and North America. If it weren't for a narrow gap in the Panamanian jungle (Darién Gap), one could travel on a single, continuous road, the Pan-American Highway, from Alaska to Patagonia. Although there had been proposals to build this missing segment, its construction never began. The missing segment would need to cross a dense jungle---a formidable engineering challenge. 

Talking about engineering challenges, the most famous thing in Panama is probably the Panama Canal. Built between 1904 and 1914, it’s one of the largest and most difficult engineering projects ever undertaken. Although eventually constructed by the U.S., other countries also discovered the enormous potential of connecting the Atlantic Ocean and the Pacific Ocean. At the end of the 19th century, the French had already made an unsuccessful attempt to build it. 

The Canal is 82 kilometers long. It has an artificial lake in the middle, located above sea level. At both ends, a sophisticated system lifts and lowers the ships. An engineering system of this complexity requires active maintenance. Initially maintained by the U.S., and after a period of joint management, Panama started to own and operate the Canal in 1999. 

According to the government-owned operator, more than 13,000 vessels passed through the Canal last year alone. In the same year, the tolls paid by the passing vessels contributed around [2 billion USD to the government’s budget](https://www.swissinfo.ch/spa/panam%C3%A1-canal_canal-de-panam%C3%A1-entrega-2.080-millones-de-d%C3%B3lares-al-fisco-como-aporte-anual/47199818#:~:text=Indic%C3%B3%20que%20%22los%20ingresos%20por,millones%20de%20d%C3%B3lares%22%20para%202022.).

## Niger

Niger is one of the poorest countries today. There are many reasons behind this, one of them being its unfortunate geography---most of the country lies in the Sahara Desert. Around the time it became independent from France, large uranium reserves were discovered in the northern part of the country. There was hope that this resource---if managed properly---could lift many Nigeriens out of poverty. 

Judged by virtually all statistics, uranium didn’t bring about the large-scale changes that many were optimistic about. Since the beginning, uranium mining has been dominated by French conglomerates. According to some, the former colonizer, France, was in a position to negotiate deals that favored its nuclear industry (but not necessarily the Nigerien people). 

The most significant mines are in the twin mining towns of Arlit and Akokan. Far from the country’s capital and located in the desert, Arlit is Niger’s 6th most populous city, and much of its economic life is centered around uranium. In this part of the world, water is a scarce resource. Unsurprisingly, local residents who buy water in 25-liter containers will feel resentful at foreign mining companies pumping water out of the [remaining aquifers]( https://africanarguments.org/2017/07/a-forgotten-community-the-little-town-in-niger-keeping-the-lights-on-in-france-uranium-arlit-areva/). Apart from worsening water scarcity, reports mention toxic waste and illnesses caused by the operations of the nearby mines. 

Despite its many health, environmental, and fairness implications, uranium mining employs many locals. When one of the region’s longest-running mines closed last year after 47 years of operation, [hundreds lost their jobs]( https://world-nuclear-news.org/Articles/Niger-uranium-mine-ends-operations-after-47-years). In a region with very limited economic opportunities, these lost jobs will be difficult to replace.

## Costa Rica

There are only a few countries in the world that don’t have an army---Costa Rica being one of them. It had a short civil war in 1948 when disputes around election results escalated into an armed conflict. After the civil war ended, the provisional president, José Figueres, abolished the army. The funds formerly allocated to the military are now used to finance public health and education programs.

Aside from abolishing the army, Figueres granted women and Costa Ricans of African origin the vote, nationalized the banking sector, and guaranteed public education for all---just to name a few of his reforms from his 18-month provisional presidency. Costa Rica’s democracy, the oldest in Central America, also dates to 1949. After his provisional presidency, Figueres served two more terms as president.

As is frequently the case, his family members also held political and diplomatic positions in subsequent administrations. Most importantly, his son, José María, followed him in his footsteps, serving as the country’s president between 1994 and 1998. In this year's elections, José María unsuccessfully contested the same position again. But unlike dictatorships ruled by dynasties, Costa Rica is a democracy. Even when a country has a small population like Costa Rica does, for a family to stay politically relevant for seven decades is noteworthy. 

## Haiti 

Haiti became independent in 1804, the first country to do so in the Caribbean. Its history ever since has been full of natural disasters and man-made suffering. For instance, in the 200 years since its independence, the country suffered 32 coups, the most recent one in 2004. 

During the 2004 political instability, a tropical storm struck Haiti, leaving thousands dead due to flooding and landslides in the northern part of the country. Four years later, another tropical storm and several hurricanes resulted in further deaths and property damage. In another two years, the already devastated population suffered its worst earthquake in over 200 years. 

On 12 January 2010, a magnitude 7.0 Mw earthquake struck the country, with an epicenter only 25 kilometers away from the capital, Port-au-Prince. Although there have been larger earthquakes globally since then, its epicenter was close to heavily populated areas, affecting around three million people. The resulting death toll was estimated to be somewhere between 100 and 200 thousand people. Property damage was also significant: many historically important buildings, such as the Presidential Palace and the Port-au-Prince Cathedral, were damaged beyond repair. The same year Haitians were trying to clean up the devastation left behind by the earthquake, the country suffered the world’s first modern large-scale cholera outbreak. 

We would be forgiven for thinking that a series of natural disasters would bring Haitian politicians together, but we would be wrong. Political turmoil continued throughout these years---among other things, the president was assassinated last year. As Haitians were dealing with the aftermaths of this event, a devastating earthquake struck the island again. 

## Syria

The most famous family in Syria is undoubtedly the al-Assad family, having led the county for the last 50 years. Hafez al-Assad was the first nationally prominent member of the family. Born in an Alawite (minority sect of Shia Muslims) family in western Syria, his ascension to power didn’t happen overnight---it required no fewer than three coups.

After a military career, he was involved in the 1963 Syrian coup d’état, when the Ba'ath Party, a movement urging the unification of the Arab world, came to power. Three years later, a power struggle broke out within the leading factions of the party, which deposed the traditional leadership and fortified al-Assad’s position.  In 1970, he ousted his last opponent, clearing the way for his regime that would last 30 years.

In the mid-80s, he had some health issues, bringing the issue of his successor to the forefront. But some of his family members had their own plans: his brother Rifaat attempted a coup against his regime. After the failed attempt, Rifaat was exiled to France until he was allowed to return to Syria last year.

The person who allowed him to return was Hafez’s second son, Bashar, the country’s current president. Ruling Syria since his father’s death in 2000, he wasn’t destined for a political career. Trained as an eye doctor in Syria and the U.K., his political involvement became clear only after his older brother, Bassel, had died in a car accident. 

## Nigeria

Nigeria is Africa’s most populous country. It’s a country characterized by regional variation---in climate, ethnicity, religion, and even legal systems. 

In many parts of the world, climate determines how people make a living. The climate of the southern part of the country is tropical, while the northern part is an arid steppe (and desert in the northeast). Traditionally, most were pastoralists in the north and farmers in the south. Where these groups meet, in the so-called Middle Belt (the country’s central region), the situation is susceptible to conflict.

This is especially the case when ethnic, linguistic, and religious differences also abound. Nigeria is a multi-ethnic state with hundreds of ethnicities. The three largest groups dominate one region of the country each: the Hausa in the north, the Yoruba in the southwest, and the Igbo in the southeast. Not only do these groups speak different languages, but their religious affiliations are also different. The northern part of the country follows Islam, while the southern part is Christian. Reflecting its past, Nigeria's legal system has elements of common law (from its colonial past), customary law (in some indigenous regions), and Sharia law (in the north). 

A country with this level of diversity is in a tricky position: it needs to maintain a common identity while allowing people to live their lives as they see fit. Nigeria is a federal country, structured in some ways like the United States. The advantage of this system is that it allows more regional variation than a unitary state. The official language is English---not one of the regional languages. If, in contrast, one of the regional languages were to become the official language, people from other linguistic groups could be alienated. 

## Central African Republic

By most statistics, the Central African Republic is one of the worst countries to live in. There is extreme poverty: the GDP per capita is around 400 dollars---or less than half the price of a new iPhone. Aside from the miserable economic conditions, its life expectancy is the lowest in the world, while its maternal mortality rate is among the highest. The majority of its population is illiterate. 

Since its independence from France in 1960, the country’s history follows a pattern that’s, unfortunately, not uncommon in the region. Civil wars, coups d’états, a small glimmer of hope for democratic change, then a devastating civil war again. Although leaders in countries like the Central African Republic don’t typically rise to power through their moral standing in the community, the country produced a particularly brutal and narcissistic dictator even by regional standards. 

Jean-Bédel Bokassa came to power in a coup in 1966. At the start of his regime, he suspended the constitution and dissolved the National Assembly. Not stopping there, he renamed the country the Central African Empire, declared himself president for life, and named himself the emperor. At that time, the country had approximately two million people---not the typical population of an empire. He spent an estimated 20 million dollars (equivalent to a third of the country’s annual budget) on his coronation ceremony. Eventually, he was ousted from power the same way he came to it---by a coup. 

 Many years after his regime had ended, he was convicted of murders, concealing corpses, embezzlement, and other charges in a widely publicized trial.


## Italy

If a country is large enough, chances are there will be conflicts between its regions. Italy, with an area of around 300,000 km2, is no different in this regard. It has a North-South divide, at least since its unification in the second half of the 19th century. The gap that existed at the unification only grew wider over time. Today, the southern part of the country, once occupied by the Kingdoms of Naples and Sicily, fares significantly worse on most economic indicators.

A particularly alarming statistic in Southern Italian provinces is the high unemployment rate. When unemployment is high, economic output tends to be lower. This is reflected in the southern provinces’ significantly lower GDP per capita. For example, out of Italy’s 20 administrative regions, the bottom five by GDP per capita are in Southern Italy. When people don’t see a bright future for themselves or their children, they leave. In 2010 alone, around [134,000 people]( https://www.theguardian.com/world/2011/oct/11/economic-exodus-emigration-southern-italy) left Southern Italy---most of them finding their new home in the north.

Those who stay despite the poverty, poor infrastructure, and joblessness, are exposed to the consequences of Mafia-type organized crime. According to a 2010 estimate, the yearly income of criminal organizations could reach a sum equivalent to [9% of Italy’s GDP]( https://www.telegraph.co.uk/news/worldnews/europe/italy/6957240/Italy-claims-finally-defeating-the-mafia.html). Aside from the damage to people’s livelihoods, these criminal groups were also responsible for several high-profile murders. One example of this was the Via D'Amelio bombing, which claimed the life of the distinguished anti-mafia judge Paolo Borsellino.

## Brazil

Brazil’s first capital city was Salvador. Two centuries later, in the middle of the 18th century, the capital was moved to Rio de Janeiro. Although both cities have access to the Atlantic Ocean, Rio de Janeiro was a quickly growing city closer to lucrative gold and diamond mines. When Brazil became independent in 1822, the idea of a new capital city was born. The idea, however, wouldn’t become reality for more than 130 years. 

Juscelino Kubitschek became Brazil’s president in 1956. During his campaign, hallmarked by the slogan ’50 years of progress in 5’, he promised to move the capital from Rio de Janeiro to a more central location in the country. The goal of this move was to develop Brazil’s interior regions, which lagged behind the more densely inhabited (and more prosperous) coastal regions. The new capital, Brasília, was to be planned and constructed from scratch. 

After a public competition, the jury selected the modernist architect Lúcio Costa's design. The city has a cross-axial design, which could be visualized as an airplane or a bird. One axis hosts the political and administrative buildings, while the other contains residential buildings. Construction took place at a record speed: in only 41 months, the city was finished. 

Initially designed for around half a million people, the population quickly outgrew this limit. The plan was that satellite cities, separated from the city by a 15-mile buffer zone, would accommodate the extra population. Unfortunately, the plan largely failed. Today, Brasília is frequently held up as an example of a failing city.

## Jordan

When a war breaks out in a country, people try to flee to a safer, typically neighboring one. Jordan has been relatively safe by regional standards for the last seven decades. This means that whenever the security situation worsened in one of its neighbors, a large wave of refugees arrived in the country.

The first wave of refugees were Palestinians who fled during the Arab-Israeli War in 1948. Followed by subsequent waves, most Palestinians received Jordanian citizenship. According to some estimates, the number of Palestinian refugees in the country is approximately [2 million]( https://www.unrwa.org/where-we-work/jordan). Around one in five live in refugee camps. 

More recently, during the Iraq War in 2003, up to 1 million Iraqis found shelter in Jordan. Most of them have since returned, with around 130 thousand remaining in the country. Since the start of the Syrian civil war, an estimated 1.4 million Syrian refugees have arrived in Jordan, with around half of them being registered refugees. Some live in two large refugee camps, but most live among the local population. 

Jordan has a population of around 10 million people. Providing help to over a million refugees is a sizeable undertaking for a country its size. Aside from the problems of unemployment, lack of access to healthcare, and poverty, the refugee population also suffers from a lack of clean water. The country, being one of the most water-scarce countries on earth, has a hard time providing water even to its citizens, let alone to a more marginalized refugee population.

## Libya

When the major European powers set their eyes on colonizing Africa at the end of the 19th century, Italy also started a search for potential colonies. Besides the Horn of Africa, today’s Libya would belong to its sphere of influence. In the early years of the 20th century, Libya was under the rule of the significantly weakened Ottoman Empire.

In 1911, Italy invaded the territory. The Italo-Turkish War, as it came to be known, lasted around a year and led to the establishment of two colonies: Italian Tripolitania and Italian Cyrenaica. Notable for its military innovations, this was the first war in which a military carried out an aerial reconnaissance operation and an aerial bombardment. To put things in perspective, the bombardment meant dropping 4 grenades of around 2 kilograms each from a very simple airplane. 

In 1923, a conflict between the Italian military and rebel forces associated with the Senussi broke out. Led by Omar al-Mukhtar, the Senussi were a political-religious order seeking to reform the lives of Bedouin tribesmen in the eastern part of the country. This conflict, sometimes called the Pacification of Libya, ended in 1932 after the capture and execution of al-Mukhtar. A bit of a misnomer, the war was anything but pacific. The Italian military used chemical weapons, deported a large part of Cyrenaica’s (the eastern part of the country) population from their land, and confined them to concentration camps. 

Libya remained an Italian colony until 1943. Conquered by the Allies in the North African Campaign, it was administered by British and French forces until its independence in 1951. 

## Mauritania

Mauritania abolished slavery in 1981, the last country to do so in the world. Despite being banned, owning slaves was only criminalized in 2007. Even today, the law is rarely enforced. Although the exact number of enslaved people in the country cannot be known, the 2018 Global Slavery Index estimated the currently enslaved population to be around [90 000 people]( https://www.globalslaveryindex.org/), or approximately 2% of the population. When we see a statistic like this, it’s important to bear in mind that the prevalence of slavery is difficult to estimate. This difficulty is multiplied when the country in question is located almost entirely in a vast desert. 

Furthermore, slavery doesn’t manifest itself in only one way. For instance, a U.S Department of State page mentions [7 forms of modern slavery]( https://www.state.gov/what-is-modern-slavery/), including forced labor, domestic servitude, and bonded labor. In Mauritania, chattel slavery is common. In this form of slavery, the enslaved person is the slave owner's property and can be, among other things, sold or rented out to others.

Besides the inhumane treatment of enslaved people, there is also a racial undertone in slavery in Mauritania. The country is multi-ethnic, with both lighter-skinned and darker-skinned ethnic groups. The former group, called Bidhan, are of Arab or Berber descent and comprise most slave owners; the latter group, called Haratin, are descendants of former sub-Saharan slaves and comprise most currently enslaved people.

Slavery is still a significant problem in Mauritania. Nonetheless, the last couple of decades saw a lot of progress on this issue. Anti-slavery organizations, such as the Initiative for the Resurgence of the Abolitionist Movement, increased the visibility of this human rights issue, frequently at significant personal cost. 

## Sri Lanka

In the first half of the 19th century, Sri Lanka was exporting large amounts of coffee. By the end of the century, its coffee plantations were largely wiped out by a disease, Hemileia vastatrix. This disease (a fungus) covers the leaves of a plant with spores or causes them to fall. It spread quickly---instead of just affecting a local plantation, the country’s entire coffee production was affected: it dropped from around 100 million pounds in 1870 to only 5 million [a mere 20 years later]( https://www.apsnet.org/edcenter/apsnetfeatures/Pages/ICPP98CoffeeRust.aspx).

While the disease was destroying coffee plants, tea (another plantation crop) was introduced into the country. A Scottish tea planter, James Taylor, set up a small tea plantation in the central part of the country (then called British Ceylon). After visiting the island, a fellow Scotsman, Thomas Lipton, started buying Ceylon tea from Taylor's plantation. His surname might ring familiar to some: he was the founder of the famous tea company Lipton. 
Today, Sri Lanka is the [fourth largest exporter of tea](https://www.statista.com/statistics/264188/production-of-tea-by-main-producing-countries-since-2006/).

Another of its important export products is cinnamon. Globally, there are only three countries (Indonesia, China, and Vietnam) that produce more cinnamon. Cinnamomum verum (translated as true cinnamon tree) is native to Sri Lanka; some claim it has a subtler flavor than other cinnamon species.

## Guyana

Guyana neighbors three countries: Brazil, Suriname, and Venezuela. It has territorial disputes with the latter two. In the largest dispute, Venezuela claims an area of around 160,000 km2 (out of Guyana’s 215,000 km2). The territory in question is called Guayana Esequiba by Venezuela, and it concerns the territory to the west of the Essequibo River. It’s controlled and administered by Guyana, but it has been a consistent source of tension between the two countries for a long time.

The origin of the conflict lies in an old territorial dispute between colonial powers. Venezuela was a Spanish colony, while Guyana was under Dutch and later British colonial rule. As is usually the case, both sides claimed the territory citing some historical and geographical reasons. After gold had been discovered in the 19th century, the dispute became more intense. In 1899, an international arbitration tribunal ruled almost fully in favor of the British, giving them most of the disputed territory. 

In the early 1960s, Venezuela challenged the validity of this arbitration. The newly independent Guyana and Venezuela tried to resolve the issue---without success. In 2013, an oil exploration vessel in disputed waters (i.e., the part of the ocean that’s adjacent to the disputed territory) was seized by the Venezuelan navy. As of today, there is still no agreement about where the border should be. One thing seems certain: when a disputed territory has significant natural resources, whether it's gold in the 19th century or oil in the 21st, reaching an agreement becomes more difficult. 


## Benin

Benin became independent from France in 1960. But the name Benin was only adopted in 1975---the new country was initially called Dahomey, named after the Kingdom of Dahomey. This kingdom was an important regional power before it came under French colonial rule in the late 19th century. It’s noteworthy---among other things---for its role in the slave trade and its all-female military unit. 

As a highly militarized kingdom, the Kingdom of Dahomey frequently conducted raids against other societies. During these raids, its fighters captured people who were then sold to European slave traders. The British passed the Slavery Abolition Act in 1833, which made slave ownership illegal in the British territories. As a result, the demand for slaves gradually disappeared. This Act and other factors contributed to the kingdom’s decline. After some territorial disputes and two wars with France, the kingdom was incorporated into French West Africa. 

Frequent wars require a large number of military-age men. High casualties in this population led to an unusual proposition---an all-female military unit. Named the Dahomey Amazons, they were recruited from free Dahomean women and captives. They went through rigorous training and made up about a third of the army.


## Namibia

The first inhabitants of Namibia were different hunter-gatherer tribes such as the San, Damara, and Nama people. Over the centuries, Bantu peoples started to arrive in the country, eventually making them a majority. At the end of the 19th century, most of the territory became a German colony under the name of German South West Africa. One exception to German rule was the port city of Walvis Bay and some offshore islands, which were under British rule.

The Herero and Nama people started a rebellion against the German colonizers in 1904, eventually leading to the Battle of Waterberg. The Germans, thanks to their superior firepower, quickly won the battle and drove the fleeing Herero into the Kalahari Desert, one of the driest spots in sub-Saharan Africa. According to some estimates, tens of thousands of Herero people died of starvation and dehydration. Many of those who survived the first phase of the genocide were then sent to the Shark Island concentration camp and succumbed to diseases, malnutrition, overwork, and abuse.  It was the first genocide of the 20th century. Eventually acknowledged as a genocide, the German government is now paying reparations to the affected communities. 

The Germans were defeated during World War I by the South African forces, thereby transferring control over Namibia to South Africa. Starting in the 1960s, the decolonization of Africa led to calls for Namibian self-determination. After armed insurgencies and a regional conflict, Namibia gained independence from South Africa in 1990.


## Ivory Coast

Félix Houphouët-Boigny was an unusual figure in African politics. Until his death in 1993, he was Ivory Coast’s president---a period encompassing 33 years. This is no small feat in Africa where leaders tend to stay in power for much shorter periods. At the time of his death, he was the longest-serving president in Africa; even on the world stage, he was only third, after the dictators of Cuba and North Korea. 

He maintained close ties with France throughout his years in power. Before he became the leader of the newly independent Ivory Coast, he had served in the French government. But his link to France was not only political but economic: he owned several high-value properties in Paris (and elsewhere). According to some sources, his wealth amounted to several billion dollars. By regional standards, Ivory Coast was a relatively stable country and had a strong economy, primarily driven by coffee and cocoa production. It’s not difficult to imagine how a president with strong executive power could financially benefit from a growing economy with little outside control. 

In 1983, the country’s political capital was moved from Abidjan to Yamoussoukro, making it the fourth capital city in a century. Born in a family of tribal chiefs in Yamoussoukro, Boigny’s decision was probably motivated by factors other than purely administrative or practical ones. He set out to develop the new capital city. Its flagship construction project was the Basilica of Our Lady of Peace, one of the largest churches worldwide.


## Ghana

Nii Amon Kotei was a Ghanaian artist most renowned for designing the coat of arms of Ghana. Commissioned by the British colonial government, his design was adopted when Ghana gained independence in 1957. Unchanged to this day, it features a shield with four quarters divided by a St. Georges Cross and some other symbols. 

Although all four quarters tell us something different about Ghana, let’s focus here on only two of them: a cocoa tree and a mine shaft. The cocoa tree represents the country’s wealth in agricultural products in general and cocoa in particular. Only second to its western neighbor (Ivory Coast) in [cocoa production]( https://www.statista.com/statistics/263855/cocoa-bean-production-worldwide-by-region/), the cocoa tree makes an important contribution to the country’s economy.

Besides agricultural production, mining (as represented by the mine shaft) is also paramount to the Ghanaian economy. Although the country mines and exports a variety of metals, gold is by far the most important one.  It's one of the country’s top export products. But not only that: Ghana’s gold production is also significant internationally, in some years being the largest in Africa.

Above the shield is a black five-pointed star, the lodestar of African freedom. Also present in Ghana’s flag (designed the same year by a different artist), it became a symbol of Pan-Africanism and anti-colonialism and served as an inspiration for other African flags.    


## Gabon

Gabon, like most African countries, has substantial ethnic diversity. The majority of its people belong to different Bantu ethnic groups, the largest of which are the Fang people. Also found in neighboring Cameroon and Equatorial Guinea, their culture is rich in folklore and oral tradition.

Some of their ceremonial objects, such as face masks and reliquary figures, are valuable collector items. For instance, a Ngil mask was sold in France for [4.2 million euros earlier this week](https://www.rfi.fr/en/africa/20220327-rare-fang-mask-sold-for-%E2%82%AC4-2-million-in-france-despite-protest-gabon). The recent sale is more of a coincidence than an everyday occurrence---these masks are difficult to find these days.

The masks were used by a secret society called Ngil that administered justice and policed negative behaviors, such as witchcraft, in Fang communities. They typically had a broad forehead, almond-shaped eyes, and a long nose. Many of them were white because they were covered with white clay (kaolin). When used during the night, these masks (as part of a full costume) could make the wearer look quite intimidating.

Around 100 years ago, the secret society was banned by the French colonialists, which led to a [gradual disappearance of Ngil masks](https://www.christies.com/lot/lot-masque-fang-ngila-fang-mask-gabon-6166310/?lid=1&from=relatedlot&intobjectid=6166310). Made from wood, it's unlikely that many of them survived to this day in pristine conditions. For example, the recently sold mask had a difficult-to-replicate trajectory before going under the hammer. It was originally collected around 1917 by a colonial governor. Lying around in an attic, it was finally discovered by his descendants a century later.

## Kyrgyzstan

Issyk-Kul is a large endorheic (i.e., without outflow) mountain lake in northeastern Kyrgyzstan. Its name means "warm lake" in the Kyrgyz language. Situated in the northern Tian Shan mountains at an elevation of 1600 meters, its name doesn't refer to the outside temperature (which can be very cold indeed) but to the fact that the lake never freezes.

The lake has significant biodiversity. It contains several fish species that aren't found elsewhere, so-called endemic species. Some of them are endangered by overfishing and predation by other fish species that had been introduced in the 1970s.

Beyond its biodiversity, Issyk-Kul has also been an attractive spot for tourism for the last couple of decades. During Kyrgyzstan's Soviet era, when domestic tourism within the Soviet Union was encouraged, the empire's elite started frequenting the lake and built lavish resorts and hotels in the town of Cholpon Ata. Although there have been new developments in recent years, much of the town's current infrastructure was built in the Soviet era. The lake's region is responsible for 70% of all tourism in Kyrgyzstan, attracting around [1 million people a year (opens a pdf)](https://www.adb.org/sites/default/files/project-document/62284/41548-kgz-dpta-v1-cholponata.pdf). Many tourists to the region are domestic travelers or from former Soviet states, such as Kazakhstan or Uzbekistan.

## Chad

Chad is a multi-ethnic country, having around 200 ethnic groups. The largest ethnic group are the Sara people, with around 30% of the population belonging to this ethnicity. They live in the southern part of Chad, to the southeast of Lake Chad, an endorheic lake, fed primarily by the Chari River. This region is the country's most fertile, and a relatively small, area with a large population. In contrast, the northern and central parts of Chad cover a large area but have a low population density. In the north, we find the Saharan zone while the central part is the semi-arid Sahel, a transition zone leading to the Sudanian savannah.

In multi-ethnic countries like Chad, it's not uncommon to see different ethnic groups struggle for power. Frequently, this struggle ends in a protracted and devastating civil war. Typically, the country's leader favors his ethnicity and commits atrocities or discriminates against other ethnicities. The other ethnic groups then feel threatened and harbor hate and resentment. When they have the chance, they try to topple this leader and settle the score. The process then starts all over again.

Chad became independent in 1960. Its first leader was a man named François Tombalbaye, an ethnic Sara. An interesting side note about this leader: a photo from 1959 shows scars on his face. Not from a war-time injury or a debilitating disease but from what is called scarification, a common tradition in many African tribes. During this procedure, a person's body is intentionally cut or burned to leave permanent scars.

During his regime, Tombalbaye banned opposition parties, and inter-ethnic relations got worse, eventually leading to a civil war. He was killed in 1975. The country's next leader was Hissène Habré, an ethnic Toubou from the north. Fifteen years later, he was deposed by one of his generals, an ethnic Zaghawa. This general ruled Chad for three decades, surviving various coup d'états and rebellions. He was killed in 2021 during a military offensive against rebel groups.

## Lesotho

Lesotho holds many records in geography. For instance, it's one of only three enclave countries in the world (the other two being Vatican City and San Marino). An enclave country's territory is fully surrounded by that of another country. In Lesotho's case, that other country is South Africa. Furthermore, Lesotho is the only country whose entire territory lies above 1000 meters. To be more precise, its lowest point is at 1400 meters, beating the second country on the list, Rwanda, by 450 meters.

A mountainous landscape makes a country difficult to conquer. This probably played a role in Lesotho's independence. After all, a small territory in the middle of a much larger one isn't likely to maintain its independence for long unless it has a serious natural advantage (e.g., a difficult terrain) over a potential invader. But the historical record shows that, even then, virtually all attempts to remain independent end in failure sooner or later. Currently, the only exception to this is Lesotho.

That said, even though Lesotho is an independent state, South Africa is hugely important for its future development. An important milestone in cooperation is the Lesotho Highlands Water Project, the largest infrastructure project between the two countries. The goal of the project is to divert water from the rivers in the Lesotho Highlands to the Gauteng region in South Africa through an elaborate system of dams and tunnels. South Africa benefits from an increased water supply, while Lesotho earns royalties and generates hydroelectricity.

## South Sudan

In 2011, the United Nations admitted South Sudan as its most recent, 193rd member. The country gained its independence from Sudan after two prolonged civil wars, each lasting around 20 years. This means that a person born in the southern part of Sudan in 1955 (the year Sudan itself became independent) had to survive around 40 years of devastating civil wars before he could celebrate his 56th birthday in an independent South Sudan.

Living that long wasn't very likely, however, if we consider that life expectancy at birth in 1960 was around 48 years in Sudan and 32 years in South Sudan, [according to the World Bank](https://data.worldbank.org/indicator/SP.DYN.LE00.IN?locations=SS-SD). (It's unclear how the World Bank calculated life expectancy in 1960 for a country that hadn't existed back then.)

A mere two years into its independence, the country descended into a new civil war that lasted 7 years. During this civil war, an estimated 400,000 people were killed, and millions were displaced internally or had to flee to neighboring countries. Ethnic conflict soared, one particularly gruesome episode of which was the 2014 Bentiu massacre. Even more recently, the country experienced widespread famine, caused both by the war and a pest outbreak. In 2020, the sides agreed to a peace deal.

## Germany

If we had traveled through what is now Germany some centuries ago, we would have found a large number of sovereign, independent states. They comprised a loose political entity called the Holy Roman Empire, which existed for around 1000 years and was finally dissolved in 1806. A decade later, in 1815, it was replaced by another federation of states, the German Confederation. A primarily military alliance, it was formed by 39 independent states, including Prussia and Austria.

Being the two most powerful states, Prussia and Austria were rivals in the quest for power. The tensions between them culminated in the Seven Weeks' War in 1866, which was won by Prussia and led to the dissolution of the Confederation. The now dominant Prussia created the North German Confederation. Shortly after, a dispute between the this newly formed entity and France escalated into the Franco-Prussian war, which induced Southern German states to join the Prussia-led Confederation.

With the Confederation now involving both northern and southern states, the unification was essentially complete: in 1871, the German Empire was born, with Prussia at the helm. The mastermind behind the rise of Prussia and the main driving force behind Germany becoming a nation-state was Otto von Bismarck. He became the first Chancellor of the Empire, holding this position until 1890.

## Seychelles

For a small country of around 100 thousand inhabitants, one wouldn't expect a lot of political turmoil. We would be wrong in this case. Since Seychelles gained its independence from the United Kingdom in 1976, it saw, among other things, a successful coup d'état, various failed ones, a one-party state, the murder of an opposition leader in London, and a peaceful transition of power.

One year into its independence, the country's elected president was deposed while attending a conference outside the country. In a virtually bloodless coup, around 200 armed men took over strategic positions on Mahé, the main island. Based on a contemporary [New York Times article](https://www.nytimes.com/1977/06/07/archives/seychelles-capital-is-calm-after-coup-insurgents-secretive-about.html) that describes the aftermath of the events, one comes away with the impression that the inhabitants' lives hadn't been upended in significant ways. For example, the article notes that tourists on the island were a bit annoyed they couldn't buy alcohol temporarily and that six British officers were arrested and flown back to Europe. After this coup, the person who served as the prime minister under the deposed president formed a new government and quickly instituted a one-party state.

Four years later, in 1981, the country saw an unsuccessful coup attempt, led by an Irish mercenary and orchestrated by South Africa. A group of 44 mercenaries flew into Mahé, under the pretext that they were rugby players on holiday. Their identity was quickly discovered when a baggage search produced an AK-47, and the coup attempt failed.

In 1991, the same president who introduced the one-party state announced a return to a multiparty system. He won several elections before voluntarily turning over his presidency to another politician from his party. In 2020, the elections were won by the opposition party that hadn't been in power since 1977.

In sum, Seychellois politics of the last couple of decades produced many rather unexpected (and occasionally bizarre) episodes.

## Malta

Maltese is one of the two official languages in Malta. It is a Semitic language, making Malta the only country in the European Union to use one. How did Maltese become Malta's official language? To answer this question, we need to take a step back and briefly consider Malta's geographic location and history.

The most widely spoken Semitic language is Arabic. The Arab world is geographically close (around 300 km) to Malta, with Tunisia to its west and Libya to its south. By the beginning of the 10th century, the Arabs conquered Sicily, located to the north of Malta, and created the Emirate of Sicily, which also included Malta. The conquerors spoke a variety of Arabic called Siculo-Arabic. This language is now extinct, but it served as the starting point for the evolution of Maltese.

In 1091, Norman forces conquered Malta, and the island's re-Christianisation began. The language went through a process of Latinisation: the formerly spoken language was written down in Latin script. It's the only Semitic language written exclusively with Latin characters, albeit it has some extra, diacritically altered letters such as ċ. Over the centuries, Italian and Sicilian had a large influence over it. As a result, its vocabulary does not only contain words of Arabic origin but also many loanwords from these two languages and, more recently, English. It has been Malta's official language since 1934, and around 97% of the population consider it their mother tongue.

## Uzbekistan

Historically, cotton has played an important role in Uzbekistan. For instance, all the emblems the country had in the last roughly 100 years featured a cotton plant. Cotton production greatly increased during the Soviet era: around 70% of the Soviet Union's entire production came from the Uzbek SSR. Although cotton production is now lower than it was in the Soviet era, Uzbekistan is still among the largest producers and exporters of cotton. Nicknamed as "white gold", cotton has undoubtedly contributed to the country's economy, but its production also brought with it some problems. Let's briefly mention three of them.

First, cotton is a water-intensive crop. The water for its irrigation was supplied in part by the Aral Sea. Furthermore, rivers feeding the Aral Sea (Amu Darya and Syr Darya) were diverted for Soviet irrigation projects. As a result of all this, the Aral Sea started shrinking in the 1960s and lost around 90% of its original size, creating the world's most recent desert (Aralkum Desert) in the dried-up basin.

Second, the environmental damage was exacerbated by the extensive use of fertilizers and other chemical products, which caused soil contamination and public health problems. For example, a defoliant called Butifos (since outlawed) was blamed by locals for [higher than usual infant mortality and birth defects](https://www.newscientist.com/article/mg12416910-800-soviet-cotton-threatens-a-regions-sea-and-its-children/).

Finally, some reports claim that cotton production in Uzbekistan is rife with human rights abuses. According to a 2020 [Bureau of International Labor Affairs report](https://www.dol.gov/agencies/ilab/reports/child-labor/list-of-goods), forced labor is still used, and some earlier reports documented child labor as well.

## Suriname

Suriname has enormous ethnic diversity for a country of only 600 thousand people. To understand why this is the case, let's briefly look at some of its history. Some areas of today's Suriname were inhabited by indigenous tribes, the largest one of which were the Arawak. When Suriname became a Dutch colony in the second half of the 17th century, new plantations could be created in the fertile Guiana plains, which required a lot of human labor. Initially, this labor was provided by slaves from sub-Saharan Africa. Many of them managed to escape from the plantations and find refuge in the adjacent rainforest. The descendants of these people are the Maroons and constitute around 22% of the Surinamese.

After the Dutch abolished slavery in 1863, the labor-intensive plantations had to find another source of labor. Contract and indentured (i.e., people working without a salary for some years) workers were brought in primarily from today's Indonesia and India. Like Suriname, Indonesia was also a Dutch colony, which made transporting people to Suriname easier. Organizing the transport of Indian workers was a bit more involved because, India being under British colonial rule, arrangements needed to be made with the British. Today around 15% of the population have Javanese (from Java in Indonesia) ancestry and around 27% have Indian ancestry. We also find people with Creole, Chinese, and Brazilian heritage among the country's residents.

To sum it up, Suriname's current ethnic composition offers a glimpse of its history: the slave trade from Africa, contract and indentured workers from Asia, and recent voluntary migration from places such as Brazil.

## Burundi

Countries in the same region tend to have many things in common. One of the aspects in which Burundi is similar to some other countries in the region (Tanzania, in particular, comes to mind) is that it has recently changed its capital city. Historically, the capital of the pre-colonial Kingdom of Burundi was Gitega, which is now the second-largest city in Burundi. Around the time Burundi gained its independence from Belgium, Bujumbura became its capital city. Now it's the country's largest and economically most important city, conveniently located on the shore of Lake Tanganyika. A couple of years ago, Burundi's then-president announced that the capital would be Gitega again.

I've always wondered why countries change their capitals. I assume sometimes there is a purely political reason, such as moving to a politically more welcoming region. Other times, changing the capital can reflect, among other things, a genuine shift in the importance of certain regions within a country.

Finally, let me mention two things briefly in which Burundi is a bit atypical in sub-Saharan Africa. First, Burundi is linguistically more homogeneous than most other countries in the region: almost everyone speaks Kirundi, while in other countries there is usually a large degree of linguistic diversity. Second, the current country's territory occupies that of the pre-colonial Kingdom of Burundi. This is rare in that most current borders in the region don't coincide with those of pre-colonial states.

## Uganda

Uganda is named after Buganda, its largest region. Buganda is in the southern part of the country, and approximately 25% of the country's 45 million residents live there---many of them in the capital, Kampala. People from Buganda are collectively called the Baganda (or Ganda), while a single person from this group is a Muganda. They speak a language called Luganda.

In yesterday's post about Kenya, I mentioned that flags may contain useful information about a country. So, I duly checked Uganda's flag. It depicts a grey crowned crane, Uganda's national bird. So far so good. Today's revelation (or reminder, rather) is that names can also contain interesting pieces of information.

Suspecting that Buganda, Luganda, and Baganda may have something to do with the country's name didn't require a lot of effort because the words are almost identical. Without knowing too much about the country's history, one would guess that the Baganda people must have played an important role in shaping Uganda's past and present.

Detecting that some words share a common root is not always as straightforward as in Uganda's case. Had the words been a bit more different from one another, I could have easily overlooked the link between the country's name and its most important region.

## Kenya

I've heard about the Maasai before, but I haven't realized they are one of the largest ethnic groups in Kenya. It turns out that Kenya's flag even features a Maasai shield. A note to my future self: checking out the symbols on a flag can give me some interesting leads. Another ethnic group I found interesting are the Kalenjin: many of the famous Kenyan middle- and long-distance runners are from this ethnic group. It would be interesting to know why.

I also learned that the largest urban slum in Africa is a place called Kibera, on the outskirts of Nairobi, Kenya's capital. My questions: (1) Why there? In other words, is there a geographic, legal, or other reason why the slum is there, or is its location just the result of a random process?; (2) Did its residents move there from a particular region in Kenya (or East Africa)?

Ngũgĩ wa Thiong'o is one of the most popular Kenyan writers. He has a short story (*The Upright Revolution: Or Why Humans Walk Upright*) that has been translated into 100 languages. Maybe it's time for me to read some Kenyan literature.

While I was reading about Kenya, I noticed many names and concepts that start with the letter K---for example, Kenya itself, or the aforementioned Kalenjin and Kibera. Is this a coincidence, or is the letter K indeed common in some or all the languages Kenyans use?
