---
title: "One day, one country"
---

# Project description

Every day I randomly pick a new country, then I read and write a bit about it. The idea is to learn something new every day about the world and get into the habit of writing regularly. As I write about a new country every day, I don't have time to research things carefully. This means there may be many inaccuracies and outright errors. Please feel free to point them out to me if you encounter any. I primarily think of these notes as a public diary in which I keep track in (almost) real-time of the things I found interesting and the questions I had as I was learning about a new country.

# Overview

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r prepare-table}
posts <- tidyr::tribble(
  ~Day, ~Country, ~Continent, ~Notes, 
  "Day 2", "Uganda", "Africa", "[See notes for Uganda](#uganda)",
  "Day 1", "Kenya", "Africa", "[See notes for Kenya](#kenya)"

)
```

```{r display-table}
knitr::kable(posts)
```

# Notes

## Uganda

Uganda is named after Buganda, its largest region. Buganda is in the southern part of the
country, and approximately 25% of the country's 45 million residents live there---many
of them in the capital, Kampala. People from Buganda are collectively called
the Baganda (or Ganda), while a single person from this group is a Muganda. They
speak a language called Luganda.

In yesterday's post about Kenya, I mentioned that flags may contain useful information about a
country. So, I duly checked Uganda's flag. It depicts a grey crowned crane,
Uganda's national bird. So far so good. Today's revelation (or reminder,
rather) is that names can also contain interesting pieces of information.

Suspecting that Buganda, Luganda, and Baganda may have something to do with the country's name didn't require a lot of effort because the words are almost identical. Without knowing
too much about the country's history, one would guess that the Baganda people must have played an important role in shaping Uganda's past and
present.

Detecting that some words share a common root is not always as straightforward as in
Uganda's case. Had the words been a bit more different from one
another, I could have easily overlooked the link between the country's name and its most important region.

## Kenya

I've heard about the Maasai before, but I haven't realized they are one of the largest ethnic groups in Kenya. It turns out that Kenya's flag even features a Maasai shield. A note to my future self: checking out the symbols on a flag can give me some interesting leads. Another ethnic group I found interesting are the Kalenjin: many of the famous Kenyan middle- and long-distance runners are from this ethnic group. It would be interesting to know why.

I also learned that the largest urban slum in Africa is a place called Kibera, on the outskirts of Nairobi, Kenya's capital. My questions: (1) Why there? In other words, is there a geographic, legal, or other reason why the slum is there, or is its location just the result of a random process?; (2) Did its residents move there from a particular region in Kenya (or East Africa)?

Ngũgĩ wa Thiong'o is one of the most popular Kenyan writers. He has a short story (*The Upright Revolution: Or Why Humans Walk Upright*) that has been translated into 100 languages. Maybe it's time for me to read some Kenyan literature.

While I was reading about Kenya, I noticed many names and concepts that start with the letter K---for example, Kenya itself, or the aforementioned Kalenjin and Kibera. Is this a coincidence, or is the letter K indeed common in some or all the languages Kenyans use?
