---
title: "One day, one country"
---

# Project description

Every day I randomly pick a new country, then I read and write a bit about it. The idea is to learn something new every day about the world and get into the habit of writing regularly. As I write about a new country every day, I don't have time to research things carefully. This means there may be many inaccuracies and outright errors. Please feel free to point them out to me if you encounter any. I primarily think of these notes as a public diary in which I keep track in (almost) real-time of the things I found interesting and the questions I had as I was learning about a new country.

# Overview

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r prepare-table}
posts <- tidyr::tribble(
  ~Day, ~Country, ~Continent, ~Notes,
  "Day 16", "Namibia", "Africa", "[See notes for Namibia](#namibia)",
  "Day 15", "Ivory Coast", "Africa", "[See notes for Ivory Coast](#ivory-coast)",
  "Day 14", "Ghana", "Africa", "[See notes for Ghana](#ghana)",
  "Day 13", "Gabon", "Africa", "[See notes for Gabon](#gabon)",
  "Day 12", "Kyrgyzstan", "Asia", "[See notes for Kyrgyzstan](#kyrgyzstan)",
  "Day 11", "Chad", "Africa", "[See notes for Chad](#chad)",
  "Day 10", "Lesotho", "Africa", "[See notes for Lesotho](#lesotho)",
  "Day 9", "South Sudan", "Africa", "[See notes for South Sudan](#south-sudan)",
  "Day 8", "Germany", "Europe", "[See notes for Germany](#germany)",
  "Day 7", "Seychelles", "Africa", "[See notes for Seychelles](#seychelles)",
  "Day 6", "Malta", "Europe", "[See notes for Malta](#malta)",
  "Day 5", "Uzbekistan", "Asia", "[See notes for Uzbekistan](#uzbekistan)",
  "Day 4", "Suriname", "South America", "[See notes for Suriname](#suriname)",
  "Day 3", "Burundi", "Africa", "[See notes for Burundi](#burundi)",
  "Day 2", "Uganda", "Africa", "[See notes for Uganda](#uganda)",
  "Day 1", "Kenya", "Africa", "[See notes for Kenya](#kenya)"

)
```

```{r display-table}
knitr::kable(posts)
```

# Notes

## Namibia

The first inhabitants of Namibia were different hunter-gatherer tribes such as the San, Damara, and Nama people. Over the centuries, Bantu peoples started to arrive in the country, eventually making them a majority. At the end of the 19th century, most of the territory became a German colony under the name of German South West Africa. One exception to German rule was the port city of Walvis Bay and some offshore islands, which were under British rule.

The Herero and Nama people started a rebellion against the German colonizers in 1904, eventually leading to the Battle of Waterberg. The Germans, thanks to their superior firepower, quickly won the battle and drove the fleeing Herero into the Kalahari Desert, one of the driest spots in sub-Saharan Africa. According to some estimates, tens of thousands of Herero people died of starvation and dehydration. Many of those who survived the first phase of the genocide were then sent to the Shark Island concentration camp and succumbed to diseases, malnutrition, overwork, and abuse.  It was the first genocide of the 20th century. Eventually acknowledged as a genocide, the German government is now paying reparations to the affected communities. 

The Germans were defeated during World War I by the South African forces, thereby transferring control over Namibia to South Africa. Starting in the 1960s, the decolonization of Africa led to calls for Namibian self-determination. After armed insurgencies and a regional conflict, Namibia gained independence from South Africa in 1990.


## Ivory Coast

Félix Houphouët-Boigny was an unusual figure in African politics. Until his death in 1993, he was Ivory Coast’s president---a period encompassing 33 years. This is no small feat in Africa where leaders tend to stay in power for much shorter periods. At the time of his death, he was the longest-serving president in Africa; even on the world stage, he was only third, after the dictators of Cuba and North Korea. 

He maintained close ties with France throughout his years in power. Before he became the leader of the newly independent Ivory Coast, he had served in the French government. But his link to France was not only political but economic: he owned several high-value properties in Paris (and elsewhere). According to some sources, his wealth amounted to several billion dollars. By regional standards, Ivory Coast was a relatively stable country and had a strong economy, primarily driven by coffee and cocoa production. It’s not difficult to imagine how a president with strong executive power could financially benefit from a growing economy with little outside control. 

In 1983, the country’s political capital was moved from Abidjan to Yamoussoukro, making it the fourth capital city in a century. Born in a family of tribal chiefs in Yamoussoukro, Boigny’s decision was probably motivated by factors other than purely administrative or practical ones. He set out to develop the new capital city. Its flagship construction project was the Basilica of Our Lady of Peace, one of the largest churches worldwide.


## Ghana

Nii Amon Kotei was a Ghanaian artist most renowned for designing the coat of arms of Ghana. Commissioned by the British colonial government, his design was adopted when Ghana gained independence in 1957. Unchanged to this day, it features a shield with four quarters divided by a St. Georges Cross and some other symbols. 

Although all four quarters tell us something different about Ghana, let’s focus here on only two of them: a cocoa tree and a mine shaft. The cocoa tree represents the country’s wealth in agricultural products in general and cocoa in particular. Only second to its western neighbor (Ivory Coast) in [cocoa production]( https://www.statista.com/statistics/263855/cocoa-bean-production-worldwide-by-region/), the cocoa tree makes an important contribution to the country’s economy.

Besides agricultural production, mining (as represented by the mine shaft) is also paramount to the Ghanaian economy. Although the country mines and exports a variety of metals, gold is by far the most important one.  It's one of the country’s top export products. But not only that: Ghana’s gold production is also significant internationally, in some years being the largest in Africa.

Above the shield is a black five-pointed star, the lodestar of African freedom. Also present in Ghana’s flag (designed the same year by a different artist), it became a symbol of Pan-Africanism and anti-colonialism and served as an inspiration for other African flags.    


## Gabon

Gabon, like most African countries, has substantial ethnic diversity. The majority of its people belong to different Bantu ethnic groups, the largest of which are the Fang people. Also found in neighboring Cameroon and Equatorial Guinea, their culture is rich in folklore and oral tradition.

Some of their ceremonial objects, such as face masks and reliquary figures, are valuable collector items. For instance, a Ngil mask was sold in France for [4.2 million euros earlier this week](https://www.rfi.fr/en/africa/20220327-rare-fang-mask-sold-for-%E2%82%AC4-2-million-in-france-despite-protest-gabon). The recent sale is more of a coincidence than an everyday occurrence---these masks are difficult to find these days.

The masks were used by a secret society called Ngil that administered justice and policed negative behaviors, such as witchcraft, in Fang communities. They typically had a broad forehead, almond-shaped eyes, and a long nose. Many of them were white because they were covered with white clay (kaolin). When used during the night, these masks (as part of a full costume) could make the wearer look quite intimidating.

Around 100 years ago, the secret society was banned by the French colonialists, which led to a [gradual disappearance of Ngil masks](https://www.christies.com/lot/lot-masque-fang-ngila-fang-mask-gabon-6166310/?lid=1&from=relatedlot&intobjectid=6166310). Made from wood, it's unlikely that many of them survived to this day in pristine conditions. For example, the recently sold mask had a difficult-to-replicate trajectory before going under the hammer. It was originally collected around 1917 by a colonial governor. Lying around in an attic, it was finally discovered by his descendants a century later.

## Kyrgyzstan

Issyk-Kul is a large endorheic (i.e., without outflow) mountain lake in northeastern Kyrgyzstan. Its name means "warm lake" in the Kyrgyz language. Situated in the northern Tian Shan mountains at an elevation of 1600 meters, its name doesn't refer to the outside temperature (which can be very cold indeed) but to the fact that the lake never freezes.

The lake has significant biodiversity. It contains several fish species that aren't found elsewhere, so-called endemic species. Some of them are endangered by overfishing and predation by other fish species that had been introduced in the 1970s.

Beyond its biodiversity, Issyk-Kul has also been an attractive spot for tourism for the last couple of decades. During Kyrgyzstan's Soviet era, when domestic tourism within the Soviet Union was encouraged, the empire's elite started frequenting the lake and built lavish resorts and hotels in the town of Cholpon Ata. Although there have been new developments in recent years, much of the town's current infrastructure was built in the Soviet era. The lake's region is responsible for 70% of all tourism in Kyrgyzstan, attracting around [1 million people a year (opens a pdf)](https://www.adb.org/sites/default/files/project-document/62284/41548-kgz-dpta-v1-cholponata.pdf). Many tourists to the region are domestic travelers or from former Soviet states, such as Kazakhstan or Uzbekistan.

## Chad

Chad is a multi-ethnic country, having around 200 ethnic groups. The largest ethnic group are the Sara people, with around 30% of the population belonging to this ethnicity. They live in the southern part of Chad, to the southeast of Lake Chad, an endorheic lake, fed primarily by the Chari River. This region is the country's most fertile, and a relatively small, area with a large population. In contrast, the northern and central parts of Chad cover a large area but have a low population density. In the north, we find the Saharan zone while the central part is the semi-arid Sahel, a transition zone leading to the Sudanian savannah.

In multi-ethnic countries like Chad, it's not uncommon to see different ethnic groups struggle for power. Frequently, this struggle ends in a protracted and devastating civil war. Typically, the country's leader favors his ethnicity and commits atrocities or discriminates against other ethnicities. The other ethnic groups then feel threatened and harbor hate and resentment. When they have the chance, they try to topple this leader and settle the score. The process then starts all over again.

Chad became independent in 1960. Its first leader was a man named François Tombalbaye, an ethnic Sara. An interesting side note about this leader: a photo from 1959 shows scars on his face. Not from a war-time injury or a debilitating disease but from what is called scarification, a common tradition in many African tribes. During this procedure, a person's body is intentionally cut or burned to leave permanent scars.

During his regime, Tombalbaye banned opposition parties, and inter-ethnic relations got worse, eventually leading to a civil war. He was killed in 1975. The country's next leader was Hissène Habré, an ethnic Toubou from the north. Fifteen years later, he was deposed by one of his generals, an ethnic Zaghawa. This general ruled Chad for three decades, surviving various coup d'états and rebellions. He was killed in 2021 during a military offensive against rebel groups.

## Lesotho

Lesotho holds many records in geography. For instance, it's one of only three enclave countries in the world (the other two being Vatican City and San Marino). An enclave country's territory is fully surrounded by that of another country. In Lesotho's case, that other country is South Africa. Furthermore, Lesotho is the only country whose entire territory lies above 1000 meters. To be more precise, its lowest point is at 1400 meters, beating the second country on the list, Rwanda, by 450 meters.

A mountainous landscape makes a country difficult to conquer. This probably played a role in Lesotho's independence. After all, a small territory in the middle of a much larger one isn't likely to maintain its independence for long unless it has a serious natural advantage (e.g., a difficult terrain) over a potential invader. But the historical record shows that, even then, virtually all attempts to remain independent end in failure sooner or later. Currently, the only exception to this is Lesotho.

That said, even though Lesotho is an independent state, South Africa is hugely important for its future development. An important milestone in cooperation is the Lesotho Highlands Water Project, the largest infrastructure project between the two countries. The goal of the project is to divert water from the rivers in the Lesotho Highlands to the Gauteng region in South Africa through an elaborate system of dams and tunnels. South Africa benefits from an increased water supply, while Lesotho earns royalties and generates hydroelectricity.

## South Sudan

In 2011, the United Nations admitted South Sudan as its most recent, 193rd member. The country gained its independence from Sudan after two prolonged civil wars, each lasting around 20 years. This means that a person born in the southern part of Sudan in 1955 (the year Sudan itself became independent) had to survive around 40 years of devastating civil wars before he could celebrate his 56th birthday in an independent South Sudan.

Living that long wasn't very likely, however, if we consider that life expectancy at birth in 1960 was around 48 years in Sudan and 32 years in South Sudan, [according to the World Bank](https://data.worldbank.org/indicator/SP.DYN.LE00.IN?locations=SS-SD). (It's unclear how the World Bank calculated life expectancy in 1960 for a country that hadn't existed back then.)

A mere two years into its independence, the country descended into a new civil war that lasted 7 years. During this civil war, an estimated 400,000 people were killed, and millions were displaced internally or had to flee to neighboring countries. Ethnic conflict soared, one particularly gruesome episode of which was the 2014 Bentiu massacre. Even more recently, the country experienced widespread famine, caused both by the war and a pest outbreak. In 2020, the sides agreed to a peace deal.

## Germany

If we had traveled through what is now Germany some centuries ago, we would have found a large number of sovereign, independent states. They comprised a loose political entity called the Holy Roman Empire, which existed for around 1000 years and was finally dissolved in 1806. A decade later, in 1815, it was replaced by another federation of states, the German Confederation. A primarily military alliance, it was formed by 39 independent states, including Prussia and Austria.

Being the two most powerful states, Prussia and Austria were rivals in the quest for power. The tensions between them culminated in the Seven Weeks' War in 1866, which was won by Prussia and led to the dissolution of the Confederation. The now dominant Prussia created the North German Confederation. Shortly after, a dispute between the this newly formed entity and France escalated into the Franco-Prussian war, which induced Southern German states to join the Prussia-led Confederation.

With the Confederation now involving both northern and southern states, the unification was essentially complete: in 1871, the German Empire was born, with Prussia at the helm. The mastermind behind the rise of Prussia and the main driving force behind Germany becoming a nation-state was Otto von Bismarck. He became the first Chancellor of the Empire, holding this position until 1890.

## Seychelles

For a small country of around 100 thousand inhabitants, one wouldn't expect a lot of political turmoil. We would be wrong in this case. Since Seychelles gained its independence from the United Kingdom in 1976, it saw, among other things, a successful coup d'état, various failed ones, a one-party state, the murder of an opposition leader in London, and a peaceful transition of power.

One year into its independence, the country's elected president was deposed while attending a conference outside the country. In a virtually bloodless coup, around 200 armed men took over strategic positions on Mahé, the main island. Based on a contemporary [New York Times article](https://www.nytimes.com/1977/06/07/archives/seychelles-capital-is-calm-after-coup-insurgents-secretive-about.html) that describes the aftermath of the events, one comes away with the impression that the inhabitants' lives hadn't been upended in significant ways. For example, the article notes that tourists on the island were a bit annoyed they couldn't buy alcohol temporarily and that six British officers were arrested and flown back to Europe. After this coup, the person who served as the prime minister under the deposed president formed a new government and quickly instituted a one-party state.

Four years later, in 1981, the country saw an unsuccessful coup attempt, led by an Irish mercenary and orchestrated by South Africa. A group of 44 mercenaries flew into Mahé, under the pretext that they were rugby players on holiday. Their identity was quickly discovered when a baggage search produced an AK-47, and the coup attempt failed.

In 1991, the same president who introduced the one-party state announced a return to a multiparty system. He won several elections before voluntarily turning over his presidency to another politician from his party. In 2020, the elections were won by the opposition party that hadn't been in power since 1977.

In sum, Seychellois politics of the last couple of decades produced many rather unexpected (and occasionally bizarre) episodes.

## Malta

Maltese is one of the two official languages in Malta. It is a Semitic language, making Malta the only country in the European Union to use one. How did Maltese become Malta's official language? To answer this question, we need to take a step back and briefly consider Malta's geographic location and history.

The most widely spoken Semitic language is Arabic. The Arab world is geographically close (around 300 km) to Malta, with Tunisia to its west and Libya to its south. By the beginning of the 10th century, the Arabs conquered Sicily, located to the north of Malta, and created the Emirate of Sicily, which also included Malta. The conquerors spoke a variety of Arabic called Siculo-Arabic. This language is now extinct, but it served as the starting point for the evolution of Maltese.

In 1091, Norman forces conquered Malta, and the island's re-Christianisation began. The language went through a process of Latinisation: the formerly spoken language was written down in Latin script. It's the only Semitic language written exclusively with Latin characters, albeit it has some extra, diacritically altered letters such as ċ. Over the centuries, Italian and Sicilian had a large influence over it. As a result, its vocabulary does not only contain words of Arabic origin but also many loanwords from these two languages and, more recently, English. It has been Malta's official language since 1934, and around 97% of the population consider it their mother tongue.

## Uzbekistan

Historically, cotton has played an important role in Uzbekistan. For instance, all the emblems the country had in the last roughly 100 years featured a cotton plant. Cotton production greatly increased during the Soviet era: around 70% of the Soviet Union's entire production came from the Uzbek SSR. Although cotton production is now lower than it was in the Soviet era, Uzbekistan is still among the largest producers and exporters of cotton. Nicknamed as "white gold", cotton has undoubtedly contributed to the country's economy, but its production also brought with it some problems. Let's briefly mention three of them.

First, cotton is a water-intensive crop. The water for its irrigation was supplied in part by the Aral Sea. Furthermore, rivers feeding the Aral Sea (Amu Darya and Syr Darya) were diverted for Soviet irrigation projects. As a result of all this, the Aral Sea started shrinking in the 1960s and lost around 90% of its original size, creating the world's most recent desert (Aralkum Desert) in the dried-up basin.

Second, the environmental damage was exacerbated by the extensive use of fertilizers and other chemical products, which caused soil contamination and public health problems. For example, a defoliant called Butifos (since outlawed) was blamed by locals for [higher than usual infant mortality and birth defects](https://www.newscientist.com/article/mg12416910-800-soviet-cotton-threatens-a-regions-sea-and-its-children/).

Finally, some reports claim that cotton production in Uzbekistan is rife with human rights abuses. According to a 2020 [Bureau of International Labor Affairs report](https://www.dol.gov/agencies/ilab/reports/child-labor/list-of-goods), forced labor is still used, and some earlier reports documented child labor as well.

## Suriname

Suriname has enormous ethnic diversity for a country of only 600 thousand people. To understand why this is the case, let's briefly look at some of its history. Some areas of today's Suriname were inhabited by indigenous tribes, the largest one of which were the Arawak. When Suriname became a Dutch colony in the second half of the 17th century, new plantations could be created in the fertile Guiana plains, which required a lot of human labor. Initially, this labor was provided by slaves from sub-Saharan Africa. Many of them managed to escape from the plantations and find refuge in the adjacent rainforest. The descendants of these people are the Maroons and constitute around 22% of the Surinamese.

After the Dutch abolished slavery in 1863, the labor-intensive plantations had to find another source of labor. Contract and indentured (i.e., people working without a salary for some years) workers were brought in primarily from today's Indonesia and India. Like Suriname, Indonesia was also a Dutch colony, which made transporting people to Suriname easier. Organizing the transport of Indian workers was a bit more involved because, India being under British colonial rule, arrangements needed to be made with the British. Today around 15% of the population have Javanese (from Java in Indonesia) ancestry and around 27% have Indian ancestry. We also find people with Creole, Chinese, and Brazilian heritage among the country's residents.

To sum it up, Suriname's current ethnic composition offers a glimpse of its history: the slave trade from Africa, contract and indentured workers from Asia, and recent voluntary migration from places such as Brazil.

## Burundi

Countries in the same region tend to have many things in common. One of the aspects in which Burundi is similar to some other countries in the region (Tanzania, in particular, comes to mind) is that it has recently changed its capital city. Historically, the capital of the pre-colonial Kingdom of Burundi was Gitega, which is now the second-largest city in Burundi. Around the time Burundi gained its independence from Belgium, Bujumbura became its capital city. Now it's the country's largest and economically most important city, conveniently located on the shore of Lake Tanganyika. A couple of years ago, Burundi's then-president announced that the capital would be Gitega again.

I've always wondered why countries change their capitals. I assume sometimes there is a purely political reason, such as moving to a politically more welcoming region. Other times, changing the capital can reflect, among other things, a genuine shift in the importance of certain regions within a country.

Finally, let me mention two things briefly in which Burundi is a bit atypical in sub-Saharan Africa. First, Burundi is linguistically more homogeneous than most other countries in the region: almost everyone speaks Kirundi, while in other countries there is usually a large degree of linguistic diversity. Second, the current country's territory occupies that of the pre-colonial Kingdom of Burundi. This is rare in that most current borders in the region don't coincide with those of pre-colonial states.

## Uganda

Uganda is named after Buganda, its largest region. Buganda is in the southern part of the country, and approximately 25% of the country's 45 million residents live there---many of them in the capital, Kampala. People from Buganda are collectively called the Baganda (or Ganda), while a single person from this group is a Muganda. They speak a language called Luganda.

In yesterday's post about Kenya, I mentioned that flags may contain useful information about a country. So, I duly checked Uganda's flag. It depicts a grey crowned crane, Uganda's national bird. So far so good. Today's revelation (or reminder, rather) is that names can also contain interesting pieces of information.

Suspecting that Buganda, Luganda, and Baganda may have something to do with the country's name didn't require a lot of effort because the words are almost identical. Without knowing too much about the country's history, one would guess that the Baganda people must have played an important role in shaping Uganda's past and present.

Detecting that some words share a common root is not always as straightforward as in Uganda's case. Had the words been a bit more different from one another, I could have easily overlooked the link between the country's name and its most important region.

## Kenya

I've heard about the Maasai before, but I haven't realized they are one of the largest ethnic groups in Kenya. It turns out that Kenya's flag even features a Maasai shield. A note to my future self: checking out the symbols on a flag can give me some interesting leads. Another ethnic group I found interesting are the Kalenjin: many of the famous Kenyan middle- and long-distance runners are from this ethnic group. It would be interesting to know why.

I also learned that the largest urban slum in Africa is a place called Kibera, on the outskirts of Nairobi, Kenya's capital. My questions: (1) Why there? In other words, is there a geographic, legal, or other reason why the slum is there, or is its location just the result of a random process?; (2) Did its residents move there from a particular region in Kenya (or East Africa)?

Ngũgĩ wa Thiong'o is one of the most popular Kenyan writers. He has a short story (*The Upright Revolution: Or Why Humans Walk Upright*) that has been translated into 100 languages. Maybe it's time for me to read some Kenyan literature.

While I was reading about Kenya, I noticed many names and concepts that start with the letter K---for example, Kenya itself, or the aforementioned Kalenjin and Kibera. Is this a coincidence, or is the letter K indeed common in some or all the languages Kenyans use?
