---
title: "One day, one country"
---

# Project description

Every day I randomly pick a new country, then I read and write a bit about it. The idea is to learn something new every day about the world and get into the habit of writing regularly. As I write about a new country every day, I don't have time to research things carefully. This means there may be many inaccuracies and outright errors. Please feel free to point them out to me if you encounter any. I primarily think of these notes as a public diary in which I keep track in (almost) real-time of the things I found interesting and the questions I had as I was learning about a new country.

# Overview

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r prepare-table}
posts <- tidyr::tribble(
  ~Day, ~Country, ~Continent, ~Notes, 
  "Day 4", "Suriname", "South America", "[See notes for Suriname](#suriname)",
  "Day 3", "Burundi", "Africa", "[See notes for Burundi](#burundi)",
  "Day 2", "Uganda", "Africa", "[See notes for Uganda](#uganda)",
  "Day 1", "Kenya", "Africa", "[See notes for Kenya](#kenya)"

)
```

```{r display-table}
knitr::kable(posts)
```

# Notes

## Suriname

Suriname has enormous ethnic diversity for a country of only 600 thousand people.  To understand why this is the case, let’s briefly look at some of its history. Some areas of today’s Suriname were inhabited by indigenous tribes, the largest one of which were the Arawak. When Suriname became a Dutch colony in the second half of the 17th century, new plantations could be created in the fertile Guiana plains, which required a lot of human labor. Initially, this labor was provided by slaves from sub-Saharan Africa. Many of them managed to escape from the plantations and find refuge in the adjacent rainforest. The descendants of these people are the Maroons and constitute around 22% of the Surinamese.

After the Dutch abolished slavery in 1863, the labor-intensive plantations had to find another source of labor. Contract and indentured (i.e., people working without a salary for some years) workers were brought in primarily from today’s Indonesia and India. Like Suriname, Indonesia was also a Dutch colony, which made transporting people to Suriname easier. Organizing the transport of Indian workers was a bit more involved because, India being under British colonial rule, arrangements needed to be made with the British. Today around 15% of the population have Javanese (from Java in Indonesia) ancestry and around 27% have Indian ancestry. We also find people with Creole, Chinese, and Brazilian heritage among the country’s residents. 

To sum it up, Suriname’s current ethnic composition offers a glimpse of its history: the slave trade from Africa, contract and indentured workers from Asia, and recent voluntary migration from places such as Brazil. 


## Burundi

Countries in the same region tend to have many things in common. One of the aspects in which Burundi is similar to some other countries in the region (Tanzania, in particular, comes to mind) is that it has recently changed its capital city. Historically, the capital of the pre-colonial Kingdom of Burundi was Gitega, which is now the second-largest city in Burundi. Around the time Burundi gained its independence from Belgium, Bujumbura became its capital city. Now it’s the country’s largest and economically most important city, conveniently located on the shore of Lake Tanganyika. A couple of years ago, Burundi’s then-president announced that the capital would be Gitega again.

I’ve always wondered why countries change their capitals. I assume sometimes there is a purely political reason, such as moving to a politically more welcoming region. Other times, changing the capital can reflect, among other things, a genuine shift in the importance of certain regions within a country.

Finally, let me mention two things briefly in which Burundi is a bit atypical in sub-Saharan Africa. First, Burundi is linguistically more homogeneous than most other countries in the region: almost everyone speaks Kirundi, while in other countries there is usually a large degree of linguistic diversity. Second, the current country’s territory occupies that of the pre-colonial Kingdom of Burundi. This is rare in that most current borders in the region don’t coincide with those of pre-colonial states. 


## Uganda

Uganda is named after Buganda, its largest region. Buganda is in the southern part of the
country, and approximately 25% of the country's 45 million residents live there---many
of them in the capital, Kampala. People from Buganda are collectively called
the Baganda (or Ganda), while a single person from this group is a Muganda. They
speak a language called Luganda.

In yesterday's post about Kenya, I mentioned that flags may contain useful information about a
country. So, I duly checked Uganda's flag. It depicts a grey crowned crane,
Uganda's national bird. So far so good. Today's revelation (or reminder,
rather) is that names can also contain interesting pieces of information.

Suspecting that Buganda, Luganda, and Baganda may have something to do with the country's name didn't require a lot of effort because the words are almost identical. Without knowing
too much about the country's history, one would guess that the Baganda people must have played an important role in shaping Uganda's past and
present.

Detecting that some words share a common root is not always as straightforward as in
Uganda's case. Had the words been a bit more different from one
another, I could have easily overlooked the link between the country's name and its most important region.

## Kenya

I've heard about the Maasai before, but I haven't realized they are one of the largest ethnic groups in Kenya. It turns out that Kenya's flag even features a Maasai shield. A note to my future self: checking out the symbols on a flag can give me some interesting leads. Another ethnic group I found interesting are the Kalenjin: many of the famous Kenyan middle- and long-distance runners are from this ethnic group. It would be interesting to know why.

I also learned that the largest urban slum in Africa is a place called Kibera, on the outskirts of Nairobi, Kenya's capital. My questions: (1) Why there? In other words, is there a geographic, legal, or other reason why the slum is there, or is its location just the result of a random process?; (2) Did its residents move there from a particular region in Kenya (or East Africa)?

Ngũgĩ wa Thiong'o is one of the most popular Kenyan writers. He has a short story (*The Upright Revolution: Or Why Humans Walk Upright*) that has been translated into 100 languages. Maybe it's time for me to read some Kenyan literature.

While I was reading about Kenya, I noticed many names and concepts that start with the letter K---for example, Kenya itself, or the aforementioned Kalenjin and Kibera. Is this a coincidence, or is the letter K indeed common in some or all the languages Kenyans use?
