---
title: "One book, (at least) one idea 4: No good men among the living by Anand Gopal"
description: |
  Black and white thinking, and what to do about it.
author:
  - name: Daniel Banki
    url: https://twitter.com/banki_daniel
date: 2022-06-06
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: false
    highlight: default
    highlight_downlit: true
creative_commons: CC BY
categories:
- book reflections
editor_options: 
  markdown: 
    wrap: sentence
preview: no-good-men-book-cover.jpg  
draft: false
---

# Introduction

Recent events made me realize how little I know about Afghanistan.
To partially remedy this, I read [No good men among the living](https://www.amazon.com/No-Good-Men-Among-Living/dp/0805091793) by Anand Gopal, a book I've had on my (virtual) shelf for a long time.
A one-sentence summary of the book is that it follows the ordeals of three Afghans (a Taliban commander, a tribal warlord, and a housewife turned politician) after the US invasion in 2001.

Although it's difficult to remember what I knew about Afghanistan before reading the book, I must have thought that the Taliban are bad, and consequently, that the Afghan warlords fighting against them must be good.
Implicitly, I assumed two things.
First, that it's always possible to tell whether someone is a Talib.
Second, that if the Taliban are bad, then the other group must be good.

This is an example of binary thinking: there are no shades of grey, everything is either black or white.
This way of thinking is common, in part because it's appropriate in some situations.
After all, some situations in life are indeed black and white.
And even when they are not, the cost of gathering extra information is, in many cases, higher than the cost of not having a nuanced view.

## Examples

## Example 1: Talib or not

Let's start with the simple-looking question of what makes someone a Talib.
US Supreme Court Justice Potter Stewart famously said the following about pornography: "I know it when I see it." I must have thought something similar about the Taliban.
Surely, they must be easily identifiable in some way, either by their ethnicity, language, or a line in their passports.
In retrospect, my prior assumption seems a bit naïve.

Some people clearly belong to the Taliban, like its leadership or its fighters who fight under the Taliban flag.
It's a lot murkier, however, when it comes to more borderline cases.
When a person's group membership cannot be easily verified, we must try to guess it based on some subjective criteria.
This raises difficulties even in theory, let alone in practice.
For instance, is a shopkeeper who helped a Taliban commander with information a Talib himself?
How about the person who fought alongside the Taliban back in the day but abandoned fighting many years ago?

In a country like Afghanistan where political volatility is high, coalitions form and dissolve, and people change their alliances frequently, many could be labeled a Talib if we used a sufficiently expansive definition.

## Example 2: Good or bad 

We are quick to jump to the conclusion that people are either good or bad, especially if don't know much about them.
If it's John, the local baker, whose character we must judge, we are less likely to fall into this mental trap.
We know he makes decent bread, but we also know that he is sometimes rude to his employees.
We know these things about him because we may have been a customer of his for the last 20 years, and we saw him in myriads of settings.
Starting from the mundane, like the way he handles customer complaints or negotiates with suppliers, to the extraordinary, like the time he was happy his daughter got into college or the time he went through a painful divorce.

For most of us, Afghanistan is not like John, but more like a distant aunt who visited us once when we were a child.
We don't remember her face, her name, or anything else about her.
We only remember that terrible face powder we accidentally tasted when we were forced to greet her with a kiss on the cheek.
We only have one piece of information about this aunt, and not a very pleasant one at that, we can use to construct a mental image of her.

Similarly, most of us haven't observed a wide range of behaviors from the different fighting factions in Afghanistan.
Consequently, if the only thing we hear about one of these factions, the Taliban, is that they are evil, it's easy to think that the Afghan warlords fighting against them are good simply because they are fighting against evil.
But history repeatedly reminds us that someone fighting against evil need not be a saint himself.

Frequently, the choice is between a bad outcome and a worse one.
The book's title (No good men among the living), a Pashtun proverb, also attests to this reality: it's more common that all sides are bad than that one side is purely good while the other is purely bad.

Let me illustrate this with a simple example.
We should be able to say that a regime that murders its dissidents is worse than the other that 'only' imprisons them.
But this doesn't mean that either of them is good.
Sometimes we fail to see how bad the alternative is, and we erroneously think that the choice is between evil and good.
We are then baffled why people tolerate the bad option.
Maybe because they could see, unlike us, that the other option wasn't very good either.

## What to do about binary thinking?

A possible antidote to binary thinking is to find a mental doppelganger---a setting like the one we are interested in, but one that we know better.

As every country had its fair share of wars and other conflicts, a natural choice would be the history of our country.
The specifics of the conflict are unlikely to be the same, but that doesn't matter: the two conflicts need to be similar only on the relevant dimension.
When it comes to our own country, we surely have our preferred narrative, but it's also more difficult to blind ourselves to the atrocities and injustices committed by our side.
We may say that, on balance, our preferred side was better.
We may say that our side's atrocities were lesser or more justifiable.
But at least we are aware of them.

Whenever we know little about a country, we don't have the buzzing voice in our heads telling us that it's not all black and white.
So, the first step to overcoming binary thinking is to remind ourselves, by thinking of a doppelganger, that judging the situation is probably not as easy as it looks.
And complicated situations call for deliberation, which, in turn, requires information.
From both sides of the conflict.

Finding a doppelganger helps us withhold judgment while we actively seek out information that goes against our preferred narrative.
Once we collected all the necessary information, we may continue to think that the situation is indeed black and white.
That's great---we certainly learned some interesting things in the process.
And in any case, the mental exercise wasn't in vain: the next situation we encounter probably won't be black and white.

```{r include=FALSE}

book_disclaimer <- paste0(here::here(), "/_disclaimers/one-book-one-idea-disclaimer.Rmd")



```

```{r disclaimer, child = book_disclaimer}





```
