---
title: "One book, (at least) one idea 1: Complications by Atul Gawande"
description: |
  A reflection about trade-offs in surgery.
author:
  - name: Daniel Banki
    url: https://twitter.com/banki_daniel
date: 09-29-2021
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: false
    highlight: default
    highlight_downlit: true
creative_commons: CC BY
categories:
- book reflections
draft: false
editor_options: 
  markdown: 
    wrap: sentence
preview: complications-book-cover.jpg    
---

In his book [*Complications: A Surgeon's Notes on an Imperfect Science*](http://atulgawande.com/book/complications/), Atul Gawande writes about some delicate ethical issues in surgery (and medicine more broadly).
It's common to find books that include the magical words 'ethical issues.' It's a lot less common (and very refreshing indeed) when one of the recurrent themes in such a book is a candid discussion of trade-offs: situations in which to get more of one thing you value, you must give up something else you value.
I focus on trade-offs in this reflection because they seem to be so rarely discussed relative to how central they are to many important decisions.
There is no easy way to categorize the types of trade-offs surgeons face; I opted to categorize them based on whether they are related to (1) who should perform the operation, (2) which treatment or procedure to use, (3) who should choose the treatment.

## Surgeon-related trade-offs

Should an operation be performed by a medical resident (i.e., a doctor yet in professional training) or an experienced doctor?
The answer may depend on the unit of analysis.
As a society, we want to have surgeons in the future.
Residents need to practice their skills to become better surgeons, for which it's unavoidable that they perform surgery.
At the same time, when I'm the one to be operated on, I want the most experienced surgeon I can find.
The next patient probably has similar desires.
The trade-off could be summarized as follows: we are very happy for residents to practice their surgical skills as long as it's on someone else.

This phenomenon is by no means unique to surgery.
Yet I suspect people feel a lot more strongly about it in surgery than in most other domains.
But then, why not talk about it more openly?

Two reasons come to mind.
First, hospital administrators/public health officials/doctors are reluctant to talk about this trade-off because this would be an admission that not all people receive the same standard of care.
When many people pay for healthcare and increasingly think of it as a service, the message of variable outcomes (in order to provide someone with a training opportunity) may not be overly popular.
Second, we as patients may not fully be aware of this trade-off: in an extreme case, we may be on an operating table, with a wide variety of doctors (some more experienced, some less) contributing to the operation.

## Procedure-related trade-offs

Another trade-off I found interesting is when surgeons must accept worse short-term outcomes (for a group of patients) to improve long-term outcomes (for another, larger group of patients).
In surgery, this would be as follows: A new procedure is available, but no one is skilled at performing it at first.
In the first X operations, surgeons are more likely to make avoidable mistakes.
As a result, patients as a group are initially worse off under the new procedure than they would have been under the old procedure.
Once surgeons gain sufficient experience, patients as a group are better off under the new procedure.
In other words, the new, superior procedure comes at the cost of worse health outcomes for the initial patients.

\
If this is known to patients ex-ante, it's difficult to imagine that many of them would rush to sign up for a new procedure (barring those who are ineligible for the old procedure).
Do we as a society miss out on improved drugs and procedures as a result?

## Patient autonomy trade-offs

In the two previously mentioned trade-offs, the individual's interest clashes with that of society.
Here, let me focus on a trade-off without this feature---patient autonomy.
Can taking away patient autonomy make the patient better off in some cases, as judged by himself?

Consider the following two situations.

Example 1: You are in a vulnerable physical, mental, or emotional state in which you cannot judge what is best for you.
You may express a preference for treatment X now, but you would go for treatment Y were it not for your vulnerable state.
What should a doctor do: go with your stated preference (X) or with what he believes to be your true preference (Y)?
In the abstract, he should act in line with your true preference (Y).
But because your true preference cannot be observed, he must make a judgment call, which implies exposing himself to two types of errors.
He can override your stated preference (and provide treatment Y), which means that he violates your right to choose if he is wrong.
Alternatively, he can follow your stated preference (X), which means that he fails to provide you with the best possible care if he is wrong.
Which type of error is less costly?

Example 2: You must make a decision between two treatment options on behalf of a loved one (e.g., a child or a spouse in a coma).
Let's say the two options are such that one would most likely leave the loved one with a small but permanent injury.
The other one allows most patients to make a full recovery, but a small number of them become paralyzed.
In a situation like this, a parent or a spouse may prefer to give up the right to choose, because they consider the alternative (i.e., the thought that their decision could potentially have contributed to a loved one's unnecessary suffering) worse.

## Conclusion

I came away from this book with a renewed appreciation for what surgeons do.
When we think of a good surgeon, we tend to think about someone with, among other things, manual dexterity and attention to detail.
An update to my mental list is in order: the ability to manage complex trade-offs seems to be a very important, yet routinely overlooked aspect of being a good surgeon.

```{r include=FALSE}

book_disclaimer <- paste0(here::here(), "/_disclaimers/one-book-one-idea-disclaimer.Rmd")



```

```{r disclaimer, child = book_disclaimer}





```
